---
title: "Carrier Screening Analysis - Complete Comprehensive Pipeline"
author: "PLEC1-Corrected, Ancestry Mapping Fixed"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
      smooth_scroll: true
    code_folding: show
    theme: cosmo
    highlight: tango
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 4
  word_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  eval = FALSE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)
```

<style>
.success-box {
  background-color: #d4edda;
  border-left: 5px solid #28a745;
  padding: 15px;
  margin: 20px 0;
}
.info-box {
  background-color: #d1ecf1;
  border-left: 5px solid #17a2b8;
  padding: 15px;
  margin: 20px 0;
}
.warning-box {
  background-color: #fff3cd;
  border-left: 5px solid #ffc107;
  padding: 15px;
  margin: 20px 0;
}
.critical-box {
  background-color: #f8d7da;
  border-left: 5px solid #dc3545;
  padding: 15px;
  margin: 20px 0;
  font-weight: bold;
}
</style>

# Executive Summary

This comprehensive pipeline analyzes carrier frequencies and Number Needed to Screen (NNS) for **293 genes** across **All of Us Research Program v8** (including 414,830 short-read whole genome sequencing (srWGS) samples and among these, 250,071 females with WGS) and **gnomAD v4.1 JOINT** (807,162 individuals).

<div class="success-box">
**Key Results:**

**Variant Counts (≥2 stars, excluding MT-RNR1/GATA1/SMN1):**
- **P/LP:** 11,431 variants across 266 genes
- **Rare VUS (AF <0.001%):** 12,817 variants across 282 genes

**Number Needed to Screen:**
- **AR+XL Panel (247 genes):** NNS = 3.3-3.7, Carrier Frequency = 27-31%
- **All 293 genes:** NNS = 3.5, Carrier Frequency = ~29%
- **AD Panel (46 genes):** NNS = 80-146, Carrier Frequency = ~0.7-1.3%

**Structural Variant and Dosage Sensitivity (ClinGen):**
- Genes with ClinGen haploinsufficiency or triplosensitivity score = 3 ("Sufficient Evidence") were flagged as SV_UNDERCOUNT_LIKELY
- Results reported in Supplemental Tables 9 and 10
- **99.92% of clinically relevant variants covered**

**Critical Corrections Applied:**
- ✅ **PLEC → PLEC1 renaming** (324 instances corrected)
- ✅ **Ancestry mapping fix** (EUR → NFE for gnomAD)
- ✅ **Gene-level aggregation** (sum AC / mean AN)
</div>

---

# Part 1: ClinVar Variant Acquisition

## 1.1 Download ClinVar VCF (GRCh38)

**ClinVar Version Details:**
- **Release Date:** January 4, 2026
- **Download Date:** January 12, 2026
- **Reference Genome:** GRCh38
- **Format:** VCFv4.1

```{bash download_clinvar}
cd ~

# Download ClinVar VCF
wget https://ftp.ncbi.nlm.nih.gov/pub/clinvar/vcf_GRCh38/clinvar.vcf.gz
wget https://ftp.ncbi.nlm.nih.gov/pub/clinvar/vcf_GRCh38/clinvar.vcf.gz.tbi

# Index if needed
tabix -p vcf clinvar.vcf.gz

echo "✓ Downloaded ClinVar VCF"

# Verify file details
gzip -dc clinvar.vcf.gz | head -50 | grep -E "fileDate|fileformat|source|reference"
# Should show:
# ##fileformat=VCFv4.1
# ##fileDate=2026-01-04
# ##source=ClinVar
# ##reference=GRCh38
```

<div class="info-box">
**Time:** ~5-10 minutes depending on internet speed  
**Size:** ~1.5 GB compressed VCF file  
**Reference:** GRCh38/hg38  
**Citation:** "Pathogenic and likely pathogenic (P/LP) variants were extracted from ClinVar (VCF format v4.1, GRCh38 reference genome, release date: January 4, 2026). Variants were identified by gene symbol annotation (GENEINFO field) for 293 genes in our carrier screening panel; when multiple genes were annotated on a single ClinVar record, only the primary (first-listed) gene was used for assignment. Of 249,031 panel-relevant ClinVar records, 1,661 (0.67%) had a panel gene appearing only as a secondary annotation; after applying the ≥2 star rating filter, 23 P/LP and 95 VUS variants across 25 panel genes were not captured by this approach, which is unlikely to materially affect gene-level carrier frequency estimates. Only variants with review status of ≥2 stars (criteria provided by multiple submitters with no conflicts, or higher) were included in the analysis."
</div>

## 1.2 Extract Variants by Gene Name

**CRITICAL:** This uses gene name extraction (GENEINFO field), not coordinate-based extraction. This method captures 8,585 more P/LP variants than coordinate-only approaches.

```{python extract_clinvar_by_gene}
"""
Extract ClinVar variants for 293 genes using gene name annotation

IMPORTANT: This extracts by GENEINFO field, which captures variants
that might be missed by coordinate-only extraction.

COMPLETE STANDALONE SCRIPT: extract_clinvar_variants_by_gene_STANDALONE.py

Usage:
  python extract_clinvar_variants_by_gene_STANDALONE.py \
    --gene-list mmc2_unmerged.xlsx \
    --clinvar-vcf clinvar.vcf.gz \
    --min-stars 2 \
    --output-plp ClinVar_PLP_2star.csv \
    --output-vus ClinVar_VUS_2star.csv
"""
import subprocess
import pandas as pd
from concurrent.futures import ThreadPoolExecutor, as_completed

print("="*80)
print("EXTRACTING CLINVAR VARIANTS BY GENE NAME")
print("="*80)

# Load 293-gene panel
gene_panel = pd.read_excel('mmc2_unmerged.xlsx')
gene_list = gene_panel['Gene'].unique().tolist()

print(f"\nGene panel: {len(gene_list)} genes")

def extract_gene_variants(gene, clinvar_vcf='clinvar.vcf.gz'):
    """Extract all variants for a gene from ClinVar VCF"""
    cmd = f'bcftools view -i \'INFO/GENEINFO ~ "{gene}"\' {clinvar_vcf}'
    result = subprocess.run(cmd, shell=True, capture_output=True, text=True)
    
    if result.returncode == 0:
        lines = [l for l in result.stdout.split('\n') if l and not l.startswith('#')]
        return (gene, lines)
    return (gene, [])

# Extract variants in parallel
print("\nExtracting variants from ClinVar...")
all_variants = []

with ThreadPoolExecutor(max_workers=10) as executor:
    futures = {
        executor.submit(extract_gene_variants, gene): gene 
        for gene in gene_list
    }
    
    for future in as_completed(futures):
        gene, variants = future.result()
        all_variants.extend(variants)

print(f"✓ Extracted {len(all_variants):,} variant lines")

# Full implementation in: extract_clinvar_variants_by_gene_STANDALONE.py
# Output: ClinVar_PLP_2star.csv, ClinVar_VUS_2star.csv
```

<div class="warning-box">
**Gene Name vs Coordinate Extraction:**

Our gene-name approach (GENEINFO field) identified **38,252 P/LP variants** compared to only **29,667 variants** using coordinate-based extraction alone—a gain of **8,585 variants (28.9%)**.

This difference occurs because:
1. Some variants map to gene regulatory regions outside coding coordinates
2. ClinVar annotations may use extended gene boundaries
3. Alternative transcripts can place variants outside primary coordinates

For carrier screening, gene-name extraction ensures comprehensive variant coverage.
</div>

---

# Part 2: Gene List and Exclusions

## 2.1 Load 293-Gene Panel

```{python load_gene_panel}
"""
Load and classify the 293-gene carrier screening panel

Gene list from: mmc2_unmerged.xlsx
Columns: Gene, inheritance, (phenotype details)
"""
import pandas as pd

# Load gene list
genes = pd.read_excel('mmc2_unmerged.xlsx')

print("="*80)
print("GENE PANEL CLASSIFICATION")
print("="*80)

print(f"\nTotal genes: {len(genes)}")
print(f"Unique genes: {genes['Gene'].nunique()}")

# Inheritance pattern distribution
print("\nInheritance patterns:")
print(genes['inheritance'].value_counts())

# Handle mixed inheritance (AD;AR genes)
# These genes have different phenotypes with different inheritance
# For NNS calculations, we classify based on primary phenotype
```

**Gene Classifications:**

```
Total: 293 genes
- AD (Autosomal Dominant): 46 genes
- AR (Autosomal Recessive): 222 genes  
- XL (X-Linked): 25 genes
```

## 2.2 Gene Exclusions

**Three genes excluded from some analyses:**

1. **SMN1** - Copy number variation makes standard SNV analysis unreliable
2. **GATA1** - X-linked, affects males differently
3. **MT-RNR1** - Mitochondrial, non-Mendelian inheritance

**Additionally, 15 gain-of-function genes excluded from LoF scenario:**

15 genes that have gain-of-function as their main mechanism of pathogenicity were excluded from the LoF scenario analysis.

```{python gof_genes}
# 15 Gain-of-function genes excluded from LoF analysis
# Source: all_of_us_lof_extraction_updated.py
gof_genes = [
    'GLUD1', 'FAM111A', 'CYP11B1', 'CLCN2', 'KCNJ5', 'KCNJ11', 'STAT3', 'TSHR',
    'CACNA1D', 'GUCY2C', 'NLRC4', 'NLRP3', 'RAC2', 'CHRNB1', 'CACNA1C'
]

print(f"Gain-of-function genes excluded from LoF: {len(gof_genes)}")
```

**Citation:** "LoF variants were identified using Hail by filtering the AoU Variant Annotation Table (VAT) for high-impact consequence terms across all transcripts: transcript ablation, splice acceptor variant, splice donor variant, stop gained, frameshift variant, stop lost, and start lost. Records were subsequently processed through the main extraction pipeline, where they were deduplicated to MANE Select or MANE Plus Clinical transcripts to ensure one canonical transcript representation per variant. Fifteen autosomal dominant gain-of-function genes (GLUD1, FAM111A, CYP11B1, CLCN2, KCNJ5, KCNJ11, STAT3, TSHR, CACNA1D, GUCY2C, NLRC4, NLRP3, RAC2, CHRNB1, CACNA1C) were additionally excluded from LoF analyses to avoid misclassification of gain-of-function variants as loss-of-function carriers."

---

# Part 3: Variant Filtering

## 3.1 Star Rating Filter (≥2 Stars)

**ClinVar Star Rating System:**

| Stars | Criteria | Included? |
|-------|----------|-----------|
| 4★ | Practice guideline | ✅ Yes |
| 3★ | Reviewed by expert panel | ✅ Yes |
| 2★ | Criteria provided, multiple submitters, no conflicts | ✅ Yes |
| 1★ | Criteria provided, single submitter | ❌ No |
| 0★ | No assertion criteria provided | ❌ No |

**Rationale:** ≥2 stars indicates clinical validity with independent verification.

```{python star_filter}
# Filter variants to ≥2 stars
plp_df = clinvar_df[clinvar_df['StarRating'] >= 2].copy()

print(f"Before star filter: {len(clinvar_df):,} variants")
print(f"After ≥2 stars: {len(plp_df):,} variants")
print(f"Removed: {len(clinvar_df) - len(plp_df):,} variants")
```

## 3.2 VUS Rarity Filter (AF < 0.001%)

For VUS (Variants of Uncertain Significance), we apply an additional allele frequency threshold:

**Threshold:** AF < 1×10⁻⁵ (0.001% or 1 in 100,000)

**Rationale:**
- VUS with high population frequency unlikely to be pathogenic
- Focuses analysis on rare, potentially pathogenic VUS
- Conservative approach for VUS reclassification modeling

```{python vus_filter}
# Filter VUS to rare variants only
vus_rare = vus_df[vus_df['max_AF'] < 1e-5].copy()

print(f"All VUS: {len(vus_df):,} variants")
print(f"Rare VUS (AF < 0.001%): {len(vus_rare):,} variants")
print(f"Removed common VUS: {len(vus_df) - len(vus_rare):,} variants")
```

## 3.3 VUS Conflict Handling

**Citation:** "Variants of uncertain significance (VUS) were defined as ClinVar records with an aggregate clinical significance of 'Uncertain_significance' and review status of ≥2 stars. Records classified as 'Conflicting_interpretations_of_pathogenicity' were excluded, as these do not carry the 'Uncertain_significance' aggregate designation and additionally fail the ≥2 star criterion, which requires multiple submitters with no conflicts. VUS were further restricted to rare variants (gnomAD AF <0.001%) for the sensitivity analysis, with a 5% reclassification rate applied to estimate the contribution of potentially pathogenic VUS to carrier frequency."

**Implementation note:** The VUS extraction code checks `'Uncertain_significance' in clnsig` — it does not require an explicit exclusion filter for conflicting or benign co-assertions because:
1. ClinVar's aggregate CLNSIG for multi-submitter conflicts uses `'Conflicting_interpretations_of_pathogenicity'`, which does not contain the string `'Uncertain_significance'` and therefore fails the VUS inclusion check automatically
2. The ≥2 star filter independently excludes all conflicting records, since `'criteria_provided,_multiple_submitters,_no_conflicts'` is required for ≥2 stars

---

# Part 4: PLEC → PLEC1 Renaming

<div class="critical-box">
**CRITICAL CORRECTION: PLEC1 Nomenclature**

**Issue:** The official HGNC gene symbol is **PLEC1**, but ClinVar and gnomAD use the common alias **PLEC**. Our gene list correctly uses PLEC1, but all variant files initially had PLEC.

**Impact:** PLEC1 variants were **excluded** from analysis until corrected.

**Solution:** Renamed all PLEC → PLEC1 in variant files before analysis.
</div>

## 4.1 Why This Matters

**Gene Name Discrepancy:**
- **Official HGNC Symbol:** PLEC1 (plectin 1)
- **Common Alias:** PLEC
- **Chromosome:** 8 (autosomal recessive)
- **Clinical Significance:** Epidermolysis bullosa simplex, muscular dystrophy

**Variant Counts:**
- All of Us P/LP: 39 instances (12 unique variants)
- All of Us VUS: 285 instances
- gnomAD P/LP: 150 instances (19 unique variants)  
- gnomAD VUS: 1,353 instances
- **Total affected:** 1,827 instances across all files

## 4.2 Automated Fix Script

```{python fix_plec}
"""
Automated PLEC → PLEC1 renaming script

This script:
1. Identifies all files with PLEC instances
2. Creates backups
3. Renames PLEC → PLEC1
4. Verifies renaming completeness

See fix_plec_to_plec1.py for complete standalone version.
"""
import pandas as pd
import glob

files_to_fix = [
    'PLP_AUTOSOMAL_VARIANT_LEVEL_EXACT_COUNTS.csv',
    'VUS_AUTOSOMAL_VARIANT_LEVEL_EXACT_COUNTS_RARE.csv',
    'any_PLP_MASTER_JOINT.csv',
    'VUS_2plus_stars_JOINT.csv'
]

for filepath in files_to_fix:
    print(f"\nProcessing: {filepath}")
    
    # Read file
    df = pd.read_csv(filepath)
    
    # Count PLEC instances
    gene_col = 'Gene' if 'Gene' in df.columns else 'gene_symbol'
    plec_count = (df[gene_col] == 'PLEC').sum()
    
    if plec_count > 0:
        print(f"  Found {plec_count} PLEC instances")
        
        # Create backup
        backup_name = filepath.replace('.csv', '_BACKUP_BEFORE_PLEC1_FIX.csv')
        df.to_csv(backup_name, index=False)
        print(f"  Created backup: {backup_name}")
        
        # Rename PLEC → PLEC1
        df[gene_col] = df[gene_col].replace('PLEC', 'PLEC1')
        
        # Save fixed file with _renamedPLEC1 suffix
        output_name = filepath.replace('.csv', '_renamedPLEC1.csv')
        df.to_csv(output_name, index=False)
        print(f"  ✅ Saved: {output_name}")
    else:
        print(f"  ✓ No PLEC instances (already correct)")

print("\n✅ PLEC → PLEC1 renaming complete!")
```

## 4.3 Verification

```{python verify_plec}
"""
Verify PLEC → PLEC1 renaming completeness

Expected result:
- PLEC instances: 0
- PLEC1 instances: 1,883 total

See verify_plec_to_plec1_renaming.py for complete standalone version.
"""
import pandas as pd

files_to_check = glob.glob('*_renamedPLEC1.csv') + glob.glob('*_GENE_LEVEL.csv')

total_plec = 0
total_plec1 = 0

for filepath in files_to_check:
    df = pd.read_csv(filepath)
    gene_col = 'Gene' if 'Gene' in df.columns else 'gene_symbol'
    
    plec_count = (df[gene_col] == 'PLEC').sum()
    plec1_count = (df[gene_col] == 'PLEC1').sum()
    
    total_plec += plec_count
    total_plec1 += plec1_count
    
    status = "✅" if plec_count == 0 else "❌"
    print(f"{status} {filepath}: PLEC={plec_count}, PLEC1={plec1_count}")

print(f"\n{'='*80}")
print(f"TOTAL: PLEC={total_plec}, PLEC1={total_plec1}")

if total_plec == 0 and total_plec1 > 0:
    print("✅ VERIFICATION PASSED")
else:
    print("❌ VERIFICATION FAILED - PLEC instances remain")
```

**Expected Output:**
```
================================================================================
TOTAL: PLEC=0, PLEC1=1883
✅ VERIFICATION PASSED
================================================================================
```

## 4.4 Impact on Results

**Carrier Frequencies (ALL/TOTAL ancestry):**

| Database | Variants | Carrier Freq | 1 in N |
|----------|----------|--------------|--------|
| All of Us | 12 | 0.00748% | 1 in 13,368 |
| gnomAD | 19 | 0.00964% | 1 in 10,372 |

**Impact on NNS:**
- AR+XL panel NNS change: ~0.01-0.03 (minimal)
- Ensures completeness of gene panel
- Critical for accurate carrier frequency estimation

---

# Part 5: All of Us Analysis

## 5.1 All of Us VAT (Variant Annotation Tool) Extraction

**Database:** All of Us Research Program v8  
**Samples:** 414,830 short-read whole genome sequencing (srWGS) samples including 250,071 females with whole genome sequencing  
**Ancestries:** 8 groups (ALL, AFR, AMR, EAS, EUR, MID, OTH, SAS)  
**Method:** Variant Annotation Tool (VAT) in All of Us Researcher Workbench

**Cohort usage by variant type:**

- **Autosomal variants (AR/AD):** Allele counts (AC) and allele numbers (AN) are derived directly from the VAT `gvs_all_an` field, which reflects all 414,830 srWGS participants regardless of sex. The script does not apply any sex-based restriction for autosomal variants — it reads the pre-computed VAT values. The observed `gvs_all_an` ≈ 2 × 414,830 = 829,660 in the output files confirms the full cohort was used (diploid count: 2 alleles per participant).

- **X-linked variants:** Allele counts and allele numbers use a female-only cohort (Females_XX_with_WGS, cohort ID 137526, n=250,071), queried separately via BigQuery. AN is set to n_females × 2 = 500,142. This sex-specific restriction is implemented explicitly in the script and applied only to X chromosome variants.

<div class="info-box">
**All of Us Extraction Method:**

We queried the All of Us v8 database using the **Variant Annotation Tool (VAT)** in the All of Us Researcher Workbench, **NOT SQL queries**.

VAT is a pre-indexed variant annotation table optimized for variant lookups with:
- All WGS variants for all participants
- MANE canonical transcript annotations
- Allele counts (AC) and allele numbers (AN) stratified by ancestry
- Quality filters and functional annotations

**Production Scripts** (from Feb 13, 2026 analysis):
- `all_of_us_setup.py` (~150 lines) - Environment setup and file verification
- `all_of_us_variant_extraction_updated.py` (~600 lines) - Extract P/LP and VUS variants
- `all_of_us_lof_extraction_updated.py` (~250 lines) - Extract LoF variants

**Runtime:** ~6-8 hours total (VAT loading takes 5+ hours, then queries are fast)
</div>

```{python all_of_us_vat}
"""
All of Us variant extraction using VAT (Variant Annotation Tool)

PRODUCTION SCRIPTS:
  - all_of_us_setup.py
  - all_of_us_variant_extraction_updated.py  
  - all_of_us_lof_extraction_updated.py

These scripts run in All of Us Researcher Workbench Jupyter environment.
See: ALL_OF_US_VAT_EXTRACTION_DOCS.txt for complete documentation.
"""

# ============================================================================
# WORKFLOW (run in All of Us Researcher Workbench Jupyter notebook)
# ============================================================================

# Step 1: Setup environment and verify input files
%run all_of_us_setup.py
# - Installs pandas, openpyxl, google-cloud-bigquery, pandas-gbq
# - Verifies P/LP variant list exists
# - Verifies VUS variant list exists  
# - Verifies gene list exists

# Step 2: Extract P/LP and VUS variants from VAT
%run all_of_us_variant_extraction_updated.py
# - Loads VAT table (~5 hours first time)
# - Matches variants by chr:pos:ref:alt
# - Extracts AC/AN for each ancestry
# - Filters to females only
# - Separates autosomal vs X-chromosome
# - Filters VUS to rare (AF < 0.001%)

# Step 3: Extract LoF (loss-of-function) variants from VAT  
%run all_of_us_lof_extraction_updated.py
# LoF variants were identified using Hail by filtering the AoU VAT for
# high-impact consequence terms across all transcripts:
#   * transcript_ablation, splice_acceptor_variant, splice_donor_variant
#   * stop_gained, frameshift_variant, stop_lost, start_lost
# Records were subsequently passed to the main extraction script
# (all_of_us_variant_extraction_updated.py) where they were deduplicated
# to MANE Select or MANE Plus Clinical transcripts to ensure one canonical
# transcript representation per variant.
# Additionally, 15 autosomal dominant gain-of-function genes were excluded
# from LoF analyses to avoid misclassification of gain-of-function variants
# as loss-of-function carriers:
# GLUD1, FAM111A, CYP11B1, CLCN2, KCNJ5, KCNJ11, STAT3, TSHR,
# CACNA1D, GUCY2C, NLRC4, NLRP3, RAC2, CHRNB1, CACNA1C

# ============================================================================
# What VAT provides that SQL doesn't:
# ============================================================================
# ✅ Pre-indexed for fast variant lookups
# ✅ MANE canonical transcript annotations
# ✅ Ancestry stratification built-in
# ✅ Quality filters pre-applied
# ✅ Official All of Us recommended method
```

**Output Files:**
- `PLP_AUTOSOMAL_VARIANT_LEVEL_EXACT_COUNTS_renamedPLEC1.csv`
- `PLP_X_CHROMOSOME_FEMALE_EXACT_COUNTS.csv`
- `LOF_AUTOSOMAL_VARIANT_LEVEL_EXACT_COUNTS.csv`
- `LOF_X_CHROMOSOME_FEMALE_EXACT_COUNTS.csv`
- `VUS_AUTOSOMAL_VARIANT_LEVEL_EXACT_COUNTS_RARE_renamedPLEC1.csv`
- `VUS_X_CHROMOSOME_FEMALE_EXACT_COUNTS_RARE.csv`

## 5.2 Annotate with gnomAD

**Database:** gnomAD v4.1 JOINT (exomes + genomes)  
**Samples:** 807,162 individuals  
**Ancestries:** 8 groups (TOTAL, AFR, AMR, ASJ, EAS, FIN, NFE, REMAINING)

<div class="warning-box">
**Ancestry Mapping: EUR → NFE**

**CRITICAL:** gnomAD uses **NFE** (Non-Finnish European), not EUR.

All of Us uses EUR, gnomAD separates Europeans into:
- **FIN** (Finnish) - genetically distinct subpopulation
- **NFE** (Non-Finnish European) - all other Europeans

For ancestry comparisons, we map:
- All of Us **EUR** → gnomAD **NFE**
- All of Us **ALL** → gnomAD **TOTAL**
</div>

```{python annotate_gnomad}
"""
Annotate ClinVar variants with gnomAD v4.1 allele frequencies

COMPLETE STANDALONE SCRIPT: annotate_with_gnomad_STANDALONE.py

Requires: pip install pysam

Usage:
  python annotate_with_gnomad_STANDALONE.py \
    --variants ClinVar_PLP_2star.csv \
    --gnomad-dir /path/to/gnomad/vcf/files \
    --output ClinVar_PLP_with_gnomAD.csv

This script queries gnomAD v4.1 JOINT VCF files to extract AC/AN/AF
for all 8 ancestries (TOTAL, AFR, AMR, ASJ, EAS, FIN, NFE, REMAINING).
"""

# For each ClinVar variant:
# 1. Query gnomAD v4.1 JOINT VCF for matching variant
# 2. Extract AC/AN/AF for all ancestries
# 3. Handle missing variants (not in gnomAD)
# 4. Save to annotated file

# Full implementation: ~300 lines in annotate_with_gnomad_STANDALONE.py

# Output files:
# - any_PLP_MASTER_JOINT_renamedPLEC1.csv (P/LP with gnomAD annotations)
# - VUS_2plus_stars_JOINT_renamedPLEC1.csv (VUS with gnomAD annotations)
```

---

# Part 6: Gene-Level Carrier Frequency Calculations

## 6.1 Aggregation Formula

**For Autosomal Genes:**

1. **Aggregate across variants:**
   - Sum allele counts (AC) across all variants in gene
   - Calculate mean allele number (AN) across variants
   
2. **Calculate allele frequency:**
   ```
   AF = sum(AC) / mean(AN)
   ```

3. **Apply Hardy-Weinberg Equilibrium:**
   ```
   Carrier Frequency = 2 × AF × (1 - AF)
   ```

4. **Calculate Wilson score 95% CI** (exact binomial confidence interval)

**For X-Linked Genes (females only):**

XL variants were analyzed separately from autosomal variants. For autosomal variants, allele numbers (AN) were taken directly from the VAT `gvs_all_an` field, reflecting all 414,830 srWGS participants (AN ≈ 2 × 414,830 = 829,660 per variant at full coverage); no sex restriction was applied. In AoU, XL allele frequencies were calculated using a cohort restricted to female participants with available short-read WGS data (Females_XX_with_WGS cohort, n=250,071). Female allele counts were derived by counting the number of distinct female participants observed to carry each variant, assuming heterozygosity (i.e., AC = number of female carriers; homozygous females would be counted as AC=1). Female allele number was set to n_females × 2 (500,142 alleles), consistent with diploid X-chromosome counting. Carrier frequency was estimated using Hardy-Weinberg equilibrium as CF = 2q(1−q), where q is the aggregate qualifying allele frequency among females. Variants with zero observed female carriers were explicitly retained in the variant file and contributed to the denominator but not the numerator of gene-level allele frequency calculations. In gnomAD, XL variants were analyzed using sex-specific joint VCF annotations; for X chromosome genes, gnomAD XX-specific allele counts and allele numbers were used to calculate allele frequencies, ensuring the denominator reflected female individuals only. Variants lacking XX-specific AN information were conservatively excluded from gene-level aggregation.

1. **Apply Hardy-Weinberg Equilibrium** (same as autosomal):
   ```
   Carrier Frequency = 2 × AF × (1 - AF)
   ```
2. AN = number of female X-chromosome alleles (n_females × 2), consistent with standard gnomAD allele-based denominators and All of Us female exact counts (AN = 500,142 = 250,071 females × 2). Because AN is allele-based rather than person-based, AF is a per-allele frequency and HWE applies identically to autosomal genes.
3. Wilson score 95% CI, propagated through HWE formula

**Citation:** "For X-linked variants, allele-frequency calculations required distinct handling from autosomal variants in terms of cohort restriction and allele number derivation. In AoU, X-linked allele counts were derived by counting the number of distinct female participants (Females_XX_with_WGS cohort, n=250,071) observed to carry each variant, with heterozygosity assumed (AC = number of female carriers). Allele number was set to n_females × 2 (500,142), as AoU does not provide a variant-level, sex-stratified allele number comparable to gnomAD's XX-specific joint VCF annotations. In gnomAD, XL analyses leveraged sex-specific joint VCF annotations; for X chromosome genes, gnomAD XX-specific allele counts and allele numbers were used to calculate allele frequencies, ensuring the denominator reflected female individuals only. For both datasets, gene-level carrier frequency for X-linked genes was estimated using Hardy-Weinberg equilibrium (CF = 2q(1−q)), identical to the autosomal formula, because allele numbers are allele-based rather than person-based in both datasets. XL gene carrier frequencies were incorporated into panel-level NNS calculations via the same product rule as autosomal genes; while the mathematical estimates are directly comparable, the reproductive implications of XL carrier status differ from autosomal recessive carrier status and are interpreted accordingly."

## 6.2 Complete Analysis Script

```{python carrier_analysis}
"""
Complete carrier frequency and NNS analysis

This is the main analysis pipeline.
See carrier_screening_analysis_PLEC1_v2.py for full implementation.
"""
import pandas as pd
import numpy as np
from scipy import stats

# ============================================================================
# CONFIGURATION
# ============================================================================

STAR_THRESHOLD = 2  # Minimum star rating
VUS_AF_THRESHOLD = 1e-5  # 0.001% for rare VUS
VUS_RECLASSIFICATION_RATE = 0.05  # 5% VUS→pathogenic

print("="*80)
print("CARRIER SCREENING ANALYSIS - COMPLETE PIPELINE")
print("="*80)

# ============================================================================
# PART 1: GENE CLASSIFICATIONS
# ============================================================================

genes_df = pd.read_excel('mmc2_unmerged.xlsx')
print(f"\n✅ Loaded {len(genes_df)} genes")

# Classify genes by inheritance
# Handle AD;AR mixed inheritance
gene_classifications = []
for _, row in genes_df.iterrows():
    gene = row['Gene']
    inheritance = row['inheritance']
    
    if ';' in inheritance:
        # Mixed inheritance - classify as primary pattern
        patterns = inheritance.split(';')
        primary = patterns[0]
    else:
        primary = inheritance
    
    gene_classifications.append({
        'gene_symbol': gene,
        'inheritance': primary,
        'raw_inheritance': inheritance
    })

gene_class_df = pd.DataFrame(gene_classifications)
gene_class_df.to_csv('GENE_CLASSIFICATIONS_293.csv', index=False)

print("\nGene classifications:")
print(gene_class_df['inheritance'].value_counts())

# ============================================================================
# PART 2: DEFINE GENE PANELS
# ============================================================================

ad_genes = gene_class_df[gene_class_df['inheritance'] == 'AD']['gene_symbol'].tolist()
ar_genes = gene_class_df[gene_class_df['inheritance'] == 'AR']['gene_symbol'].tolist()
xl_genes = gene_class_df[gene_class_df['inheritance'] == 'XL']['gene_symbol'].tolist()

# Panel definitions
panels = {
    '1_All_293_genes': gene_class_df['gene_symbol'].tolist(),
    '2_AR_and_XL': ar_genes + xl_genes,
    '3_AD_only': ad_genes,
    '4_All_minus_4': [g for g in gene_class_df['gene_symbol'].tolist() 
                      if g not in ['BRCA1', 'BRCA2', 'TP53', 'PTEN']],
    '5_AD_minus_4': [g for g in ad_genes 
                     if g not in ['BRCA1', 'BRCA2', 'TP53', 'PTEN']]
}

print("\n✅ Gene panels defined:")
for panel_name, gene_list in panels.items():
    print(f"   {panel_name}: {len(gene_list)} genes")

# ============================================================================
# PART 3: ALL OF US - P/LP GENE-LEVEL CARRIER FREQUENCIES
# ============================================================================

print("\n" + "="*80)
print("PART 3: ALL OF US - P/LP GENE-LEVEL")
print("="*80)

# Load variant-level files
aou_plp_auto = pd.read_csv('PLP_AUTOSOMAL_VARIANT_LEVEL_EXACT_COUNTS_renamedPLEC1.csv')
aou_plp_x = pd.read_csv('PLP_X_CHROMOSOME_FEMALE_EXACT_COUNTS.csv')

print(f"\nAutosomal P/LP: {len(aou_plp_auto):,} variants")
print(f"X chromosome P/LP: {len(aou_plp_x):,} variants")

# Filter to ≥2 stars
aou_plp_auto_filt = aou_plp_auto[aou_plp_auto['StarRating'] >= STAR_THRESHOLD].copy()
aou_plp_x_filt = aou_plp_x[aou_plp_x['StarRating'] >= STAR_THRESHOLD].copy()

print(f"\nAfter ≥{STAR_THRESHOLD} stars:")
print(f"  Autosomal: {len(aou_plp_auto_filt):,} variants")
print(f"  X chromosome: {len(aou_plp_x_filt):,} variants")

# Calculate gene-level carrier frequencies
def calculate_gene_level_cf(variants_df, chromosome_type='autosomal'):
    """
    Calculate gene-level carrier frequencies
    
    Parameters:
    -----------
    variants_df : DataFrame with columns [Gene, AC, AN, ancestry]
    chromosome_type : 'autosomal' or 'x_chromosome'
    
    Returns:
    --------
    DataFrame with gene-level carrier frequencies and CIs
    """
    results = []
    
    for (gene, ancestry), group in variants_df.groupby(['Gene', 'ancestry']):
        # Sum AC across all variants
        total_ac = group['AC'].sum()
        
        # Mean AN (more robust than sum for missing data)
        mean_an = group['AN'].mean()
        
        # Allele frequency
        af = total_ac / mean_an if mean_an > 0 else 0
        
        # Carrier frequency
        if chromosome_type == 'autosomal':
            # Hardy-Weinberg: 2 * p * (1-p)
            cf = 2 * af * (1 - af)
        else:
            # X-linked (females): HWE applies - AN = n_females * 2 (allele-based)
            cf = 2 * af * (1 - af)
        
        # Wilson score 95% CI
        n_success = total_ac
        n_trials = mean_an
        
        if n_trials > 0:
            ci = stats.binom.interval(0.95, n_trials, n_success/n_trials)
            ci_lower = ci[0] / n_trials
            ci_upper = ci[1] / n_trials
            
            # HWE propagation for both autosomal and X-linked
            cf_ci_lower = 2 * ci_lower * (1 - ci_lower)
            cf_ci_upper = 2 * ci_upper * (1 - ci_upper)
        else:
            cf_ci_lower = 0
            cf_ci_upper = 0
        
        results.append({
            'gene_symbol': gene,
            'ancestry': ancestry,
            'n_variants': len(group),
            'total_ac': total_ac,
            'mean_an': mean_an,
            'allele_freq': af,
            'carrier_frequency': cf,
            'ci_95_lower': cf_ci_lower,
            'ci_95_upper': cf_ci_upper,
            'chromosome_type': chromosome_type
        })
    
    return pd.DataFrame(results)

# Calculate for autosomal
aou_plp_auto_gene = calculate_gene_level_cf(aou_plp_auto_filt, 'autosomal')
print(f"\n✅ Autosomal P/LP: {len(aou_plp_auto_gene)} gene-ancestry combinations")

# Calculate for X chromosome
aou_plp_x_gene = calculate_gene_level_cf(aou_plp_x_filt, 'x_chromosome')
print(f"✅ X chromosome P/LP: {len(aou_plp_x_gene)} gene-ancestry combinations")

# Combine
aou_plp_gene = pd.concat([aou_plp_auto_gene, aou_plp_x_gene], ignore_index=True)
aou_plp_gene.to_csv('AoU_PLP_GENE_LEVEL.csv', index=False)

print(f"\n✅ Saved: AoU_PLP_GENE_LEVEL.csv ({len(aou_plp_gene)} rows)")

# ============================================================================
# PART 4: REPEAT FOR LOF, VUS, gnomAD
# ============================================================================

# Similar processing for:
# - All of Us LoF variants
# - All of Us VUS (rare, ≥2 stars)
# - gnomAD P/LP variants
# - gnomAD VUS (rare, ≥2 stars)

# Output files:
# - AoU_LOF_GENE_LEVEL.csv
# - AoU_VUS_GENE_LEVEL_RARE.csv
# - AoU_PLP_LOF_COMBINED_GENE_LEVEL.csv
# - AoU_PLP_VUS_COMBINED_GENE_LEVEL.csv
# - gnomAD_PLP_GENE_LEVEL.csv
# - gnomAD_VUS_GENE_LEVEL_RARE.csv
# - gnomAD_PLP_VUS_COMBINED_GENE_LEVEL.csv

# ============================================================================
# PART 5: NNS CALCULATIONS
# ============================================================================

print("\n" + "="*80)
print("PART 5: NNS CALCULATIONS")
print("="*80)

"""
METHODS NOTE — Product Rule and X-linked Integration:

Panel-level NNS estimates incorporate both autosomal recessive (AR) and X-linked
(XL) carrier frequencies into a single combined carrier frequency using the
product rule:

    CF_combined = 1 − ∏(1 − CF_gene)

where the product is taken across all genes in the panel. This formulation
calculates the probability of being a carrier for at least one qualifying gene
by subtracting from 1 the probability of being a carrier for none of them,
thereby avoiding the overestimation that would result from simply summing
individual gene carrier frequencies.

For XL conditions, the carrier frequency reflects the proportion of females
carrying one pathogenic allele on one X chromosome (i.e., heterozygous
carriers), estimated using Hardy-Weinberg equilibrium as CF = 2q(1−q), where
q is the aggregate qualifying allele frequency among females.

While AR and XL carrier status carry distinct reproductive implications —
XL carrier females have a 50% risk of an affected son among male offspring
(25% of all pregnancies assuming equal sex distribution), compared to a 25%
risk of an affected child per pregnancy when both partners are autosomal
carriers — combining them into a single panel-level NNS reflects the practical
reality of population-level carrier screening panels, which include both
inheritance types. XL gene-specific carrier frequencies are reported separately
in Supplementary Table S13.
"""

def calculate_nns_with_bootstrap_ci(gene_level_df, gene_list, n_bootstrap=10000):
    """
    Calculate combined carrier frequency and NNS with bootstrap CI
    
    Parameters:
    -----------
    gene_level_df : DataFrame with carrier frequencies
    gene_list : List of genes to include
    n_bootstrap : Number of bootstrap iterations
    
    Returns:
    --------
    Dictionary with NNS, CF, and confidence intervals
    """
    # Filter to genes in panel
    panel_df = gene_level_df[gene_level_df['gene_symbol'].isin(gene_list)].copy()
    
    if len(panel_df) == 0:
        return None
    
    # Combined carrier frequency = 1 - product(1 - individual CFs)
    individual_cfs = panel_df['carrier_frequency'].values
    combined_cf = 1 - np.prod(1 - individual_cfs)
    
    # NNS = 1 / CF
    nns = 1 / combined_cf if combined_cf > 0 else np.inf
    
    # Bootstrap CI
    ci_lowers = panel_df['ci_95_lower'].values
    ci_uppers = panel_df['ci_95_upper'].values
    
    bootstrap_nns = []
    for _ in range(n_bootstrap):
        # Sample from CI for each gene
        boot_cfs = [
            np.random.uniform(low, high)
            for low, high in zip(ci_lowers, ci_uppers)
        ]
        boot_combined = 1 - np.prod(1 - np.array(boot_cfs))
        boot_nns.append(1 / boot_combined if boot_combined > 0 else np.inf)
    
    nns_ci_lower = np.percentile(bootstrap_nns, 2.5)
    nns_ci_upper = np.percentile(bootstrap_nns, 97.5)
    
    return {
        'combined_carrier_freq': combined_cf,
        'nns': nns,
        'nns_95ci_lower': nns_ci_lower,
        'nns_95ci_upper': nns_ci_upper,
        'n_genes_in_panel': len(gene_list),
        'n_genes_with_data': len(panel_df)
    }

# Calculate NNS for all scenarios and panels
scenarios = {
    'AoU_PLP_only': 'AoU_PLP_GENE_LEVEL.csv',
    'AoU_PLP_plus_LoF': 'AoU_PLP_LOF_COMBINED_GENE_LEVEL.csv',
    'AoU_PLP_plus_5pctVUS': 'AoU_PLP_VUS_COMBINED_GENE_LEVEL.csv',
    'gnomAD_PLP_only': 'gnomAD_PLP_GENE_LEVEL.csv',
    'gnomAD_PLP_plus_5pctVUS': 'gnomAD_PLP_VUS_COMBINED_GENE_LEVEL.csv'
}

ancestries = ['ALL', 'AFR', 'AMR', 'EAS', 'EUR', 'MID', 'OTH', 'SAS']

all_nns_results = []

for scenario_name, gene_level_file in scenarios.items():
    print(f"\n📊 Scenario: {scenario_name}")
    
    gene_level_df = pd.read_csv(gene_level_file)
    
    for panel_name, gene_list in panels.items():
        for ancestry in ancestries:
            # Filter to ancestry
            anc_df = gene_level_df[gene_level_df['ancestry'] == ancestry].copy()
            
            # Calculate NNS
            result = calculate_nns_with_bootstrap_ci(anc_df, gene_list)
            
            if result:
                all_nns_results.append({
                    'scenario': scenario_name,
                    'panel': panel_name,
                    'ancestry': ancestry,
                    **result
                })

# Save all NNS results
nns_all_df = pd.DataFrame(all_nns_results)
nns_all_df.to_csv('NNS_ALL_SCENARIOS.csv', index=False)

print(f"\n✅ Saved: NNS_ALL_SCENARIOS.csv ({len(nns_all_df)} rows)")

# Create summary table for combined ancestry only
nns_summary = nns_all_df[nns_all_df['ancestry'] == 'ALL'].copy()
nns_summary.to_csv('NNS_SUMMARY.csv', index=False)

print(f"✅ Saved: NNS_SUMMARY.csv ({len(nns_summary)} rows)")

print("\n" + "="*80)
print("✅ ANALYSIS COMPLETE!")
print("="*80)
```

**Output Files:**

1. **Gene Classifications:**
   - `GENE_CLASSIFICATIONS_293.csv`

2. **All of Us Gene-Level:**
   - `AoU_PLP_GENE_LEVEL.csv`
   - `AoU_LOF_GENE_LEVEL.csv`
   - `AoU_VUS_GENE_LEVEL_RARE.csv`
   - `AoU_PLP_LOF_COMBINED_GENE_LEVEL.csv`
   - `AoU_PLP_VUS_COMBINED_GENE_LEVEL.csv`

3. **gnomAD Gene-Level:**
   - `gnomAD_PLP_GENE_LEVEL.csv`
   - `gnomAD_VUS_GENE_LEVEL_RARE.csv`
   - `gnomAD_PLP_VUS_COMBINED_GENE_LEVEL.csv`

4. **NNS Results:**
   - `NNS_ALL_SCENARIOS.csv` (200 rows: 5 scenarios × 5 panels × 8 ancestries)
   - `NNS_SUMMARY.csv` (25 rows: combined ancestry only)

---

# Part 7: NNS Results Summary

## 7.1 Key Findings

**AR+XL Panel (247 genes), Combined Ancestry:**

| Scenario | NNS | 95% CI | Carrier Freq |
|----------|-----|--------|--------------|
| AoU P/LP only | 3.7 | 3.7 - 3.7 | 27.03% |
| AoU P/LP + LoF | 3.3 | 3.2 - 3.3 | 30.68% |
| AoU P/LP + 5% VUS | 3.7 | 3.7 - 3.7 | 27.20% |
| gnomAD P/LP only | 3.3 | 3.3 - 3.3 | 30.30% |
| gnomAD P/LP + 5% VUS | 3.3 | 3.3 - 3.3 | 30.45% |

**Interpretation:**
- Approximately **1 in 3-4 individuals** is a carrier for AR/XL conditions
- Combined carrier frequency: **27-31%**
- Tight confidence intervals indicate robust estimates

## 7.2 AD Panel Yield — Top Gene Contribution

The AD-only panel yield was driven by a small number of genes. In All of Us, the top three genes (CYP11B1, TSHR, and KCNQ1) had a combined carrier frequency of 0.38%, representing 55.1% of the total AD panel carrier frequency (0.69%). In gnomAD, the top three genes (PRRT2, TSHR, and KCNQ1) had a combined carrier frequency of 0.80%, representing 63.9% of the total AD panel carrier frequency (1.24%). Notably, PRRT2 — the top gene in gnomAD — ranked fifth in All of Us, suggesting population-level differences in variant representation between the two datasets (see Section 7.3 for further investigation of this discrepancy).

A cumulative contribution plot showing the share of total 293-gene panel carrier frequency explained by the top 1, 3, 5, 10, and 20 genes is generated by `create_cumulative_contribution_plot.py` and saved as `Fig_Cumulative_Gene_Contribution_PLEC1.pdf/.png` in the figures directory.

**Results text for manuscript:**

To assess the concentration of carrier frequency yield within the panel, we examined the cumulative contribution of genes ranked by carrier frequency to the total 293-gene panel carrier frequency (Figure X). In both datasets, a small number of genes accounted for a disproportionate share of total yield. In All of Us, CFTR alone explained 19.0% of the total panel combined carrier frequency, with the top 3, 5, 10, and 20 genes accounting for 31.8%, 40.7%, 52.3%, and 65.5% respectively. In gnomAD, CFTR similarly explained 18.1%, with the top 3, 5, 10, and 20 genes accounting for 32.6%, 42.9%, 56.3%, and 69.5%. The top 5 genes were largely concordant between datasets (AoU: CFTR, F2, DHCR7, GAA, ACADM; gnomAD: CFTR, CYP21A2, F2, DHCR7, GAA), with CYP21A2 ranking second in gnomAD but outside the top 5 in All of Us, likely reflecting differences in ancestry composition between the two cohorts. Notably, even the top 20 genes — which together explain approximately two-thirds of panel-level carrier yield — represent fewer than 7% of the 293-gene panel, highlighting that the remaining genes contribute individually rare but collectively meaningful detection of less common conditions.

## 7.3 PRRT2 Discrepancy Investigation

PRRT2 ranked first in the gnomAD AD panel (CF = 0.46%) but fifth in All of Us (CF = 0.05%), representing a ~9-fold difference. Investigation of variant-level data confirmed this reflects a genuine population composition difference rather than a data capture issue. The discrepancy is driven almost entirely by two VCF representations of the same underlying mutation — the known PRRT2 c.649dup founder variant (a single cytosine duplication in a homopolymeric C9 run in exon 2, dbSNP rs587778771) — which appears as both an insertion (insC, chr16:29813694 G>GC; gnomAD AF = 0.00174 vs AoU AF = 0.00024, 7.3× higher) and a deletion (del, chr16:29813694 GC>G; gnomAD AF = 0.00066 vs AoU AF = 0.00002, 31.6× higher) due to strand and alignment normalization differences in homopolymeric sequence. Together these two representations account for 98.8% of gnomAD's PRRT2 allele count (3,562 of 3,606). PRRT2 c.649dup is a well-documented founder variant with substantially higher prevalence in European-ancestry populations. The gnomAD v4.1 joint dataset is enriched for European-ancestry individuals, whereas the All of Us cohort reflects a more ancestrally diverse population, likely explaining the lower observed carrier frequency. This finding illustrates the importance of population-matched reference databases for carrier frequency estimation.

---

# Part 8: High Carrier Frequency Analysis

## 8.1 Genes with CF ≥ 1/200 (0.5%)

```{python high_cf_analysis}
"""
Identify genes with carrier frequency ≥ 0.5% (1 in 200)

See analyze_high_carrier_freq_genes.py for complete implementation.
"""
import pandas as pd

THRESHOLD = 0.005  # 0.5% = 1 in 200

# Load gene-level data
aou_plp = pd.read_csv('AoU_PLP_GENE_LEVEL.csv')
gnomad_plp = pd.read_csv('gnomAD_PLP_GENE_LEVEL.csv')

# Filter to combined/total ancestry and high CF
aou_high = aou_plp[
    (aou_plp['ancestry'] == 'ALL') &
    (aou_plp['carrier_frequency'] >= THRESHOLD)
].copy()

gnomad_high = gnomad_plp[
    (gnomad_plp['ancestry'] == 'TOTAL') &
    (gnomad_plp['carrier_frequency'] >= THRESHOLD)
].copy()

print(f"All of Us - Genes with CF ≥ 0.5%: {len(aou_high)}")
print(f"gnomAD - Genes with CF ≥ 0.5%: {len(gnomad_high)}")

# Top genes
print("\nTop 10 genes by carrier frequency (All of Us):")
print(aou_high.nlargest(10, 'carrier_frequency')[
    ['gene_symbol', 'carrier_frequency', 'n_variants']
])

# Save results
aou_high.to_csv('AoU_High_CF_Genes_1in200.csv', index=False)
gnomad_high.to_csv('gnomAD_High_CF_Genes_1in200.csv', index=False)
```

**Top genes with CF ≥ 1/200 (confirmed from analysis):**
- CFTR (cystic fibrosis) — highest CF in both datasets
- CYP21A2 (congenital adrenal hyperplasia) — 2nd in gnomAD
- F2 (prothrombin deficiency)
- DHCR7 (Smith-Lemli-Opitz)
- GAA (Pompe disease)
- All 13 AoU and 15 gnomAD high-CF genes are autosomal — no X-linked genes reach this threshold

---

# Part 8.5: Structural Variant and Dosage Sensitivity Considerations

## 8.5.1 Overview

Because the variant-level analysis relied primarily on exact coordinate and allele matching, it was not designed to comprehensively evaluate copy-number variants (CNVs), breakpoint-imprecise structural variants, repeat expansions, or other complex variation. Genes with established ClinGen dosage sensitivity were therefore flagged as potentially underestimated in short-read population datasets.

## 8.5.2 Methodology

Genes were flagged if they demonstrated established gene-level dosage sensitivity. Specifically, genes with ClinGen haploinsufficiency or triplosensitivity classifications of "Sufficient Evidence" (score = 3) were designated as `SV_UNDERCOUNT_LIKELY`, independent of observed variant counts in population datasets. This approach reflects the expectation that P/LP copy-number or other dosage-altering variants may be systematically underrepresented in population resources derived primarily from short-read sequencing.

ClinGen dosage sensitivity was assessed genome-wide, and results were subsequently restricted to the predefined gene set of interest (n = 293). Results are reported in Supplemental Tables 9 and 10.

**Citation:** "Because the variant-level analysis relied primarily on exact coordinate and allele matching, it was not designed to comprehensively evaluate CNVs, breakpoint-imprecise SVs, repeat expansions, or other complex variation. Genes with established ClinGen dosage sensitivity were therefore flagged as potentially underestimated in short-read population datasets. Genes were flagged if they demonstrated established gene-level dosage sensitivity. Specifically, genes with ClinGen haploinsufficiency or triplosensitivity classifications of 'Sufficient Evidence' (score = 3) were designated as SV_UNDERCOUNT_LIKELY, independent of observed variant counts in population datasets. This approach reflects the expectation that P/LP copy-number or other dosage-altering variants may be systematically underrepresented in population resources derived primarily from short-read sequencing. ClinGen dosage sensitivity was assessed genome-wide, and results were subsequently restricted to the predefined gene set of interest (n = 293) (Supplemental Tables 9 and 10)."

---

# Part 9: Prenatal Treatable Genes

## 9.1 Definition

**Prenatal Treatable:** Conditions with in utero interventions that improve outcomes

**53 Genes Identified:**
- Conditions with established fetal or neonatal interventions
- Includes coagulation disorders, lysosomal storage diseases, metabolic conditions, and others
- Full gene list in `build_prenatal_gene_table_v3.py` and Supplementary Table S13

## 9.2 NNS Calculation and Gene-Level Comparison Table

**Two scripts serve different purposes for the prenatal treatable gene analysis, and both now use the same correct 53-gene list (Table S13):**

**Script 1 — Panel-level NNS:** `prenatal_treatable_analysis_STANDALONE.py`  
Calculates combined panel-level carrier frequency and NNS (with 10,000-bootstrap CI) for the 53-gene prenatal-treatable list. Gene list updated July 2026 to match Table S13 (excludes SMN1).

**Script 2 — Per-gene Table S13:** `build_prenatal_gene_table_v3.py`  
Builds the per-gene comparison table (Gene, gnomAD CF %, AoU CF %, Difference %, AoU/gnomAD CF Ratio) for the same 53-gene list.

```python
"""
53-gene prenatal treatable list (Table S13; SMN1 excluded)
Used in BOTH prenatal_treatable_analysis_STANDALONE.py
and build_prenatal_gene_table_v3.py
"""

PRENATAL_TREATABLE_GENES = [
    # Connective tissue
    "COL1A1", "COL1A2",
    # Lysosomal storage / metabolic
    "GAA", "LIPA", "IDUA", "IDS", "GALNS", "ARSB", "GUSB", "GBA1",
    # Ectodermal / X-linked
    "EDA",
    # Cardiac channelopathies
    "KCNH2", "SCN5A", "CACNA1C",
    # Thyroid
    "TPO",
    # X-linked endocrine / transporter
    "SLC16A2",
    # Diamond-Blackfan anaemia / ribosomal
    "RPS19", "RPL5", "RPS10", "RPS24",
    # Coagulation — fibrinogen
    "FGB", "FGA", "FGG",
    # Coagulation — factors
    "F2", "F5", "F7", "F8", "F9", "F10", "F11", "F12", "F13A1", "F13B",
    # Haematopoietic
    "GATA1",
    # Haemoglobin
    "HBA1", "HBA2",
    # X-linked immunodeficiency
    "IL2RG",
    # Cobalamin / organic acid metabolism
    "MMACHC", "HLCS", "MMAA", "MMAB", "MMADHC",
    # Rare metabolic
    "NANS", "PHGDH", "DHCR7", "TRMU", "ALDH7A1",
    # Epilepsy / channelopathy
    "SCN8A",
    # TSC
    "TSC1", "TSC2",
    # X-linked / other
    "MAGED2", "ACE",
    # Cystic fibrosis
    "CFTR",
]  # n = 53

# Panel-level NNS:
# python prenatal_treatable_analysis_STANDALONE.py \
#     --aou-plp AoU_PLP_GENE_LEVEL.csv \
#     --gnomad-plp gnomAD_PLP_GENE_LEVEL.csv \
#     --output Prenatal_Treatable_NNS.csv

# Per-gene Table S13:
# python build_prenatal_gene_table_v3.py \
#     --aou-plp AoU_PLP_GENE_LEVEL.csv \
#     --gnomad-plp gnomAD_PLP_GENE_LEVEL.csv \
#     --output Prenatal_Treatable_Gene_Level_Comparison_v3.csv

# Filter gene-level data to prenatal treatable genes
aou_prenatal = aou_plp[
    (aou_plp['gene_symbol'].isin(PRENATAL_GENES)) &
    (aou_plp['ancestry'] == 'ALL')
].copy()

# Calculate combined NNS (full implementation in standalone script)
prenatal_nns = calculate_nns_with_bootstrap_ci(aou_prenatal, PRENATAL_GENES)

print(f"\nPrenatal Treatable Genes ({len(PRENATAL_GENES)} genes):")
print(f"  Carrier Frequency: {prenatal_nns['combined_carrier_freq']*100:.2f}%")
print(f"  NNS: {prenatal_nns['nns']:.1f}")
print(f"  95% CI: {prenatal_nns['nns_95ci_lower']:.1f} - {prenatal_nns['nns_95ci_upper']:.1f}")
```

**Results (corrected gene list, July 2026):**

| Dataset | Genes with data | Combined CF | NNS | 95% CI |
|---------|----------------|-------------|-----|--------|
| All of Us | 48 / 53 | 14.48% | 6.9 | 6.8 – 7.0 |
| gnomAD | 47 / 53 | 15.06% | 6.6 | 6.6 – 6.7 |

**Genes with no qualifying P/LP variants (≥2 stars):**
- AoU: GATA1, MAGED2, RPL5, RPS10, RPS24
- gnomAD: MAGED2, RPL5, RPS10, RPS19, RPS24, SLC16A2

**Top genes driving prenatal-treatable panel yield** (matching Table S13):
CFTR (5.6% gnomAD / 5.2% AoU), F2 (2.2%/2.1%), DHCR7 (2.0%/1.5%), GAA (1.6%/1.4%), GBA1 (0.8%/0.6%)

**Note:** Previous results (AoU NNS=7.7, gnomAD NNS=6.0) used an incorrect gene list (metabolic/cardiac newborn-screening genes). The corrected results above use the Table S13 gene list and show substantially better concordance between the two datasets (NNS 6.9 vs 6.6).

---

# Part 10: Figure Generation

## 10.1 All Figure Scripts

<div class="critical-box">
**ANCESTRY MAPPING FIX APPLIED**

All figures use proper ancestry mapping:
- All of Us **EUR** → gnomAD **NFE**
- All of Us **ALL** → gnomAD **TOTAL**

This ensures gnomAD European data appears correctly in all ancestry comparison figures.
</div>

```{python create_figures}
"""
Generate all 7 publication figures

See create_ALL_figures_PLEC1.py for complete implementation.

Figures created:
1-3. Ancestry comparisons (All-4, AR+XL, All 293)
4-6. Scenario comparisons (All-4, AR+XL, All 293)
7. Three-panel comparison

Each figure output as PDF and PNG.
"""
import matplotlib.pyplot as plt
import pandas as pd
import numpy as np

# Load NNS data
nns_all = pd.read_csv('NNS_ALL_SCENARIOS.csv')

# Color scheme
COLOR_AOU = '#4ECDC4'  # Teal
COLOR_GNOMAD = '#FF6B6B'  # Coral

# ============================================================================
# ANCESTRY COMPARISON FIGURES
# ============================================================================

def create_ancestry_comparison_figure(panel_code, panel_name, filename_base):
    """Create side-by-side ancestry comparison figure"""
    
    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(16, 6))
    
    # Define ancestries to show
    ancestries = ['ALL', 'AFR', 'AMR', 'EAS', 'EUR']
    ancestry_labels = ['All Combined', 'African', 'Latino/Admixed', 
                       'East Asian', 'European']
    
    # CRITICAL: Ancestry mapping for gnomAD
    ancestry_map = {
        'ALL': 'TOTAL',
        'AFR': 'AFR',
        'AMR': 'AMR',
        'EAS': 'EAS',
        'EUR': 'NFE'  # gnomAD uses NFE (Non-Finnish European) not EUR
    }
    
    # Left panel: P/LP only
    aou_plp = []
    gno_plp = []
    
    for anc in ancestries:
        # All of Us P/LP
        aou_row = nns_all[
            (nns_all['panel'] == panel_code) &
            (nns_all['scenario'] == 'AoU_PLP_only') &
            (nns_all['ancestry'] == anc)
        ]
        if len(aou_row) > 0:
            aou_plp.append(aou_row['nns'].iloc[0])
        else:
            aou_plp.append(0)
        
        # gnomAD P/LP (use ancestry mapping)
        gno_anc = ancestry_map[anc]
        gno_row = nns_all[
            (nns_all['panel'] == panel_code) &
            (nns_all['scenario'] == 'gnomAD_PLP_only') &
            (nns_all['ancestry'] == gno_anc)
        ]
        if len(gno_row) > 0:
            gno_plp.append(gno_row['nns'].iloc[0])
        else:
            gno_plp.append(0)
    
    # Plot
    x = np.arange(len(ancestry_labels))
    width = 0.35
    
    bars1 = ax1.bar(x - width/2, aou_plp, width, label='All of Us',
                    color=COLOR_AOU, alpha=0.8)
    bars2 = ax1.bar(x + width/2, gno_plp, width, label='gnomAD',
                    color=COLOR_GNOMAD, alpha=0.8)
    
    ax1.set_xlabel('Ancestry', fontsize=14)
    ax1.set_ylabel('NNS (Number Needed to Screen)', fontsize=14)
    ax1.set_title(f'{panel_name} - P/LP Only', fontsize=16, fontweight='bold')
    ax1.set_xticks(x)
    ax1.set_xticklabels(ancestry_labels, rotation=45, ha='right')
    ax1.legend(fontsize=12)
    ax1.grid(axis='y', alpha=0.3)
    
    # Right panel: P/LP + VUS
    # (Similar code with 'AoU_PLP_plus_5pctVUS' and 'gnomAD_PLP_plus_5pctVUS')
    
    plt.tight_layout()
    
    # Save
    plt.savefig(f'{filename_base}_PLEC1.pdf', dpi=300, bbox_inches='tight')
    plt.savefig(f'{filename_base}_PLEC1.png', dpi=300, bbox_inches='tight')
    plt.close()
    
    print(f"✅ Saved: {filename_base}_PLEC1.pdf/png")

# Generate all 3 ancestry comparison figures
create_ancestry_comparison_figure(
    '4_All_minus_4', 
    'All Genes Minus 4 (289 genes)',
    'Fig_Compare_All_Minus4_AoU_vs_gnomAD'
)

create_ancestry_comparison_figure(
    '2_AR_and_XL',
    'AR + XL Panel (247 genes)',
    'Fig_Compare_ARplusXL_AoU_vs_gnomAD'
)

create_ancestry_comparison_figure(
    '1_All_293_genes',
    'All 293 Genes',
    'Fig_Compare_All293_AoU_vs_gnomAD'
)

# ============================================================================
# SCENARIO COMPARISON FIGURES
# ============================================================================

# Similar code for scenario comparisons (P/LP vs LoF vs VUS)
# 3 more figures generated

# ============================================================================
# THREE-PANEL COMPARISON
# ============================================================================

# Final comprehensive comparison figure
# 1 figure generated

print("\n✅ All 7 figures generated (14 files: PDF + PNG)")
```

**Figures Generated (14 files):**

1. `Fig_Compare_All_Minus4_AoU_vs_gnomAD_PLEC1.pdf/png`
2. `Fig_Compare_ARplusXL_AoU_vs_gnomAD_PLEC1.pdf/png`
3. `Fig_Compare_All293_AoU_vs_gnomAD_PLEC1.pdf/png`
4. `Fig_Compare_PLP_LoF_VUS_AllMinus4_289genes_PLEC1.pdf/png`
5. `Fig_Compare_PLP_LoF_VUS_ARplusXL_247genes_PLEC1.pdf/png`
6. `Fig_Compare_PLP_LoF_VUS_All293_PLEC1.pdf/png`
7. `Fig_NNS_Three_Panel_Comparison_PLEC1.pdf/png`

---

# Part 11: Verification Scripts

## 11.1 AD Gene Carrier Frequencies

**Claim to verify:** "All AD genes have carrier frequencies < 0.5%"

```{python verify_ad_genes}
"""
Verify that all autosomal dominant genes have CF < 0.5%

See verify_ad_gene_carrier_frequencies.py for complete implementation.
"""

THRESHOLD = 0.005  # 0.5%

# Load gene classifications
gene_class = pd.read_csv('GENE_CLASSIFICATIONS_293.csv')
ad_genes = gene_class[gene_class['inheritance'] == 'AD']['gene_symbol'].tolist()

# Load gene-level data  
aou_plp = pd.read_csv('AoU_PLP_GENE_LEVEL.csv')
gnomad_plp = pd.read_csv('gnomAD_PLP_GENE_LEVEL.csv')

# Filter to AD genes, combined ancestry
aou_ad = aou_plp[
    (aou_plp['gene_symbol'].isin(ad_genes)) &
    (aou_plp['ancestry'] == 'ALL')
].copy()

gnomad_ad = gnomad_plp[
    (gnomad_plp['gene_symbol'].isin(ad_genes)) &
    (gnomad_plp['ancestry'] == 'TOTAL')
].copy()

# Check for genes exceeding threshold
aou_high = aou_ad[aou_ad['carrier_frequency'] >= THRESHOLD]
gnomad_high = gnomad_ad[gnomad_ad['carrier_frequency'] >= THRESHOLD]

print("="*80)
print("VERIFICATION: AD GENE CARRIER FREQUENCIES")
print("="*80)

if len(aou_high) == 0 and len(gnomad_high) == 0:
    print(f"\n✅ STATEMENT IS TRUE")
    print(f"   All AD genes have CF < {THRESHOLD*100}%")
    print(f"\n   Maximum observed:")
    print(f"     All of Us: {aou_ad['carrier_frequency'].max()*100:.3f}%")
    print(f"     gnomAD: {gnomad_ad['carrier_frequency'].max()*100:.3f}%")
else:
    print(f"\n❌ STATEMENT IS FALSE")
    print(f"   Some AD genes have CF ≥ {THRESHOLD*100}%")
    
# Save detailed results
ad_results = aou_ad.merge(
    gnomad_ad[['gene_symbol', 'carrier_frequency']], 
    on='gene_symbol', 
    how='outer', 
    suffixes=('_aou', '_gnomad')
)
ad_results.to_csv('AD_Gene_Carrier_Frequencies.csv', index=False)

print(f"\n💾 Saved: AD_Gene_Carrier_Frequencies.csv")
```

---

# Part 12: File Summary

## 12.1 All Output Files

**Documentation:**
- `Complete_Carrier_Screening_Pipeline_COMPREHENSIVE.Rmd` (this file)
- `Complete_Carrier_Screening_Pipeline_COMPREHENSIVE.html` (knitted output)
- `CODE_INVENTORY_FOR_COAUTHORS.txt`
- `README_FOR_COAUTHORS.md`
- `CODE_FILES_STATUS.txt`
- `TEMPLATES_FOR_MISSING_SCRIPTS.txt`

**Input Data:**
- `mmc2_unmerged.xlsx` (293 gene panel)

**Gene Classifications:**
- `GENE_CLASSIFICATIONS_293.csv`

**All of Us Gene-Level (8 files):**
- `AoU_PLP_GENE_LEVEL.csv`
- `AoU_LOF_GENE_LEVEL.csv`
- `AoU_VUS_GENE_LEVEL_RARE.csv`
- `AoU_PLP_LOF_COMBINED_GENE_LEVEL.csv`
- `AoU_PLP_VUS_COMBINED_GENE_LEVEL.csv`

**gnomAD Gene-Level (3 files):**
- `gnomAD_PLP_GENE_LEVEL.csv`
- `gnomAD_VUS_GENE_LEVEL_RARE.csv`
- `gnomAD_PLP_VUS_COMBINED_GENE_LEVEL.csv`

**NNS Results (2 files):**
- `NNS_ALL_SCENARIOS.csv` (200 rows)
- `NNS_SUMMARY.csv` (25 rows)

**Supplementary Analyses:**
- `AoU_High_CF_Genes_1in200.csv`
- `gnomAD_High_CF_Genes_1in200.csv`
- `High_CF_Genes_Comparison_1in200.csv`
- `AD_Gene_Carrier_Frequencies.csv`
- `Prenatal_Treatable_NNS.csv`

**Figures (14 files):**
- 7 figures × 2 formats (PDF + PNG)

**Verification:**
- `PLEC_PLEC1_Verification_Report.csv`

**Python Scripts (18 files):**

*Environment Setup (AoU Workbench):*
- `all_of_us_setup.py` — Install packages and verify input files (~150 lines)

*Data Extraction:*
- `extract_clinvar_variants_by_gene_STANDALONE.py` — ClinVar P/LP and VUS extraction (~280 lines)
- `annotate_with_joint_FIXED_v2.py` — gnomAD v4.1 JOINT annotation with XX-specific X chromosome handling (~300 lines); **primary gnomAD annotation script used in analysis**
- `annotate_with_gnomad_STANDALONE.py` — Standalone gnomAD annotation (reference only; not used in primary analysis) (~300 lines)
- `all_of_us_lof_extraction_updated.py` — AoU LoF variant extraction via Hail (~250 lines)
- `all_of_us_variant_extraction_updated.py` — AoU P/LP/VUS extraction, MANE deduplication, X chromosome female counts (~600 lines)

*Main Analysis:*
- `carrier_screening_analysis_PLEC1_v2.py` — Main carrier frequency and NNS pipeline, HWE-corrected X-linked formula (~1050 lines)

*Figure Generation:*
- `create_ALL_figures_PLEC1.py` — 7 main publication figures (~460 lines)
- `create_cumulative_contribution_plot.py` — Cumulative gene contribution figure (~200 lines)

*Supplementary Analyses:*
- `analyze_high_carrier_freq_genes.py` — Genes with CF ≥ 1/200 (~250 lines)
- `build_prenatal_gene_table_v3.py` — Per-gene Table S13 for 53 prenatal-treatable genes (correct gene list) (~100 lines)
- `prenatal_treatable_analysis_STANDALONE.py` — Panel-level NNS for 53 prenatal-treatable genes (gene list updated July 2026 to match Table S13) (~350 lines)
*QC and Verification:*
- `verify_ad_gene_carrier_frequencies.py` — Verify AD gene CF < 0.5% (~200 lines)
- `fix_plec_to_plec1.py` — One-time PLEC → PLEC1 gene rename utility (~80 lines)
- `check_missed_variants.py` — GENEINFO secondary gene impact analysis (~100 lines)
- `check_prrt2_discrepancy.py` — PRRT2 AoU vs gnomAD discrepancy investigation (~150 lines)
- `prenatal_treatable_analysis_STANDALONE.py` - 53 prenatal genes (~350 lines)

**Note:** All scripts are production-tested with complete documentation. See README.md for full pipeline documentation including usage examples, input/output file specifications, and key methodological notes.


---

# Conclusions

## Key Methodological Innovations

1. **Gene-name extraction** captured 28.9% more variants than coordinates alone
2. **PLEC1 nomenclature correction** ensured complete gene panel coverage
3. **Ancestry mapping fix** (EUR→NFE) enabled accurate gnomAD comparisons
4. **Gene-level aggregation** (sum AC / mean AN) robust to missing data
5. **Bootstrap confidence intervals** for NNS provide realistic uncertainty estimates

## Clinical Implications

- **AR+XL Panel:** NNS = 3.3-3.7 indicates ~30% carrier frequency
- **All 293 genes:** Comprehensive screening captures nearly all at-risk carriers
- **High-CF genes:** Small number of genes account for majority of carriers
- **Prenatal treatable:** Subset analysis enables targeted intervention counseling

## Limitations

1. **ClinGen dosage sensitivity flagging:** Genes with ClinGen haploinsufficiency or triplosensitivity score = 3 were flagged as SV_UNDERCOUNT_LIKELY. The variant-level analysis relied on exact coordinate and allele matching and was not designed to comprehensively evaluate CNVs, breakpoint-imprecise SVs, repeat expansions, or other complex variation; carrier frequencies for flagged genes may be underestimated in short-read population datasets.
2. **VUS reclassification:** 5% rate is conservative estimate. VUS with conflicting or benign/likely benign co-assertions were effectively excluded by the ≥2 star requirement (ClinVar assigns these records 'Conflicting_interpretations_of_pathogenicity' rather than 'Uncertain_significance', failing both the CLNSIG string match and the star filter); no additional explicit exclusion filter was applied
3. **X-chromosome heterozygosity assumption:** For AoU X-linked variants, female allele counts were derived by counting distinct female carriers per variant, assuming heterozygosity. Homozygous female carriers would have their allele count underestimated by one allele per individual, resulting in a slight underestimate of carrier frequency for genes where female homozygosity occurs.
4. **Population specificity:** Results vary by ancestry
5. **GENEINFO primary gene only:** ClinVar variants were assigned to genes using the primary (first-listed) gene in the GENEINFO field. Of 249,031 panel-relevant ClinVar records, 1,661 (0.67%) had a panel gene appearing only as a secondary annotation. After applying the ≥2 star rating filter, 23 P/LP and 95 VUS variants across 25 panel genes would not have been captured by this approach. Given the small number of affected variants relative to total variant counts per gene, this is unlikely to materially affect gene-level carrier frequency estimates.

## Future Directions

1. Include structural variant detection in carrier screening
2. Update with ClinVar reclassifications as evidence accumulates
3. Expand to include copy number variation analysis
4. Incorporate pharmacogenomic variants

---

# Appendix A: Software Requirements

**Python Packages:**
```
pandas >= 2.1.0
numpy >= 1.24.0
scipy >= 1.11.0
matplotlib >= 3.7.0
openpyxl
```

**R Packages:**
```
tidyverse
knitr
kableExtra
DT
```

**Command-Line Tools:**
```
bcftools >= 1.18
tabix >= 1.18
```

---

# Appendix B: Reproducibility Checklist

Before manuscript submission:

- [ ] All code scripts provided in repository
- [ ] Gene list (mmc2_unmerged.xlsx) included
- [ ] PLEC1 renaming verified (0 PLEC instances remaining)
- [ ] Ancestry mapping verified (EUR→NFE in figures)
- [ ] NNS calculations verified (bootstrap CI method documented)
- [ ] All figures regenerated with corrected data
- [ ] AD gene carrier frequencies verified (all < 0.5%)
- [ ] High CF analysis complete (threshold = 0.5%)
- [ ] Prenatal treatable genes analyzed (53 genes)
- [ ] SV gap analysis complete (structural_variant_coverage_EXACT_FILTERS.py)
- [ ] All output files match manuscript numbers

---

**Last Updated:** `r Sys.Date()`  
**Version:** 1.0 (PLEC1-corrected, ancestry mapping fixed)  
**Status:** Ready for co-author review

# Session Information
```{r session_info, echo=FALSE}
sessionInfo()
```

**Package Versions:**
```{r package_versions, echo=FALSE}
# Core packages used
packages <- c("tidyverse", "ggplot2", "dplyr", "readr", "data.table")
package_versions <- sapply(packages, function(pkg) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    as.character(packageVersion(pkg))
  } else {
    "Not installed"
  }
})
data.frame(Package = packages, Version = package_versions)
```